noam_data_exploration_v1

### 2024-06-01 noam_data_exploration_v3.Rmd
Incomplete
Just need to do this because I can't figure out why I can't push so I'm trying this.

# Want to plot mean distance to centroid, but not sure how to do that with so many factors/interactions...
bray_matrix <- vegdist(species_matrix, method = "bray")
year_beta_disper <- vegan::betadisper(bray_matrix, wide_transect_level_subset$Year)
year_beta_disper #

# plot of dispersion test
plot(year_beta_disper, sub = "") # lol
boxplot(year_beta_disper, main = "Year", xlab = "")

site_beta_disper <- vegan::betadisper(bray_matrix, wide_transect_level_subset$Site)
plot(site_beta_disper, sub = "") # site five stands out
boxplot(site_beta_disper, main = "Site", xlab = "")

habitat_beta_disper <- vegan::betadisper(bray_matrix, wide_transect_level_subset$Habitat)
plot(habitat_beta_disper, sub = "") 
boxplot(habitat_beta_disper, main = "Habitat", xlab = "") # greatest variance in fringing it seems (as expected)

site_habitat_beta_disper <- vegan::betadisper(bray_matrix, wide_transect_level_subset$Site_habitat)
plot(site_habitat_beta_disper, sub = "") 
boxplot(site_habitat_beta_disper, main = "Site_habitat", xlab = "")

@@

```{r variance vs mean diversity plot, fig.height = 8, fig.width = 10}
(diversity_mean_cv_plot = 
  ggplot(data = diversity_stability) +
  geom_point(aes(x = Shannon_mean, y = Shannon_cv, colour = Site, shape = Habitat), size = 8, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +
  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )
# More diverse regions are also those that fluctuate the least in their diversity
```

@@

>>>>

```{r beta dispersion, fig.height = 15, fig.width=20}
# Want to plot mean distance to centroid, but not sure how to do that with so many factors/interactions...
# bray_matrix <- vegdist(species_matrix, method = "bray")
# year_beta_disper <- vegan::betadisper(bray_matrix, wide_transect_level_subset$Year)
# year_beta_disper #

# plot of dispersion test
# plot(year_beta_disper, sub = "") # lol
# boxplot(year_beta_disper, main = "Year", xlab = "") 
# 
# site_beta_disper <- vegan::betadisper(bray_matrix, wide_transect_level_subset$Site)
# plot(site_beta_disper, sub = "") # site five stands out
# boxplot(site_beta_disper, main = "Site", xlab = "") 
# 
# 
# habitat_beta_disper <- vegan::betadisper(bray_matrix, wide_transect_level_subset$Habitat)
# plot(habitat_beta_disper, sub = "") 
# boxplot(habitat_beta_disper, main = "Habitat", xlab = "") # greatest variance in fringing it seems (as expected)
# 
# site_habitat_beta_disper <- vegan::betadisper(bray_matrix, wide_transect_level_subset$Site_habitat)
# plot(site_habitat_beta_disper, sub = "") 
# boxplot(site_habitat_beta_disper, main = "Site_habitat", xlab = "") 
# 
stratum_beta_disper <- vegan::betadisper(bray_matrix, wide_transect_level_subset$Stratum)
# plot(site_habitat_beta_disper, sub = "") 
boxplot(stratum_beta_disper, main = "Site_habitat", xlab = "")
stratum_distances <- as.data.frame(stratum_beta_disper$distances)
colnames(stratum_distances)[1] <- "Distance"
stratum_distances$Stratum <- wide_transect_level_subset$Stratum
stratum_distances$Transect <- wide_transect_level_subset$Transect
stratum_distances$Year <- wide_transect_level_subset$Year
stratum_distances$Site <- wide_transect_level_subset$Site
stratum_distances$Habitat <- wide_transect_level_subset$Habitat
stratum_distances$Site_habitat <- wide_transect_level_subset$Site_habitat

beta_disper_means <- stratum_distances %>%
  dplyr::group_by(Year, Site, Habitat, Site_habitat) %>%
  summarise(Distance_mean = mean(Distance), 
            Distance_se = sd(Distance)/sqrt(n()))

(Distance_plot <-
  ggplot(data = beta_disper_means) +
  geom_point(aes(x = Year, y = Distance_mean, colour = Site, shape = Habitat, group = Site_habitat), size = 8, stroke = 2) +
  geom_line(aes(x = Year, y = Distance_mean, colour = Site, group = Site_habitat), size = 1.2) +
  geom_errorbar(aes(x = Year, ymin = Distance_mean - Distance_se , ymax = Distance_mean + Distance_se, colour = Site,
                    group = Site_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = site_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Habitat)+
  labs(title = "Distance to centroid\n", y = "Distance\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )

@@

```{r variance vs mean diversity plot, fig.height = 8, fig.width = 10}
(diversity_mean_cv_plot = 
  ggplot(data = diversity_stability) +
  geom_point(aes(x = Shannon_mean, y = Shannon_cv, colour = Site, shape = Habitat), size = 8, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +
  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )
# More diverse regions are also those that fluctuate the least in their diversity

(diversity_mean_cv_plot_2 = 
  ggplot(data = diversity_stability) +
  geom_point(aes(x = Shannon_mean, y = Shannon_cv, colour = Site, shape = Habitat), size = 8, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
#  facet_wrap(~Site) +
  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         ) # awesome relationship
```

@@

```{r turnover, fig.height = 30, fig.width=25}
library(codyn)
transect_level$Site_habitat <-  paste(transect_level$Site, transect_level$Habitat)
transect_level$Stratum <- paste(transect_level$Site, "-", transect_level$Habitat, "-", transect_level$Transect)
total_turnover<-
  turnover(
  df = transect_level,
  time.var = "Year",
  species.var = "Taxa",
  abundance.var = "Mean_cover",
  replicate.var = "Stratum",
  metric = "total"
)

total_turnover<- 
  total_turnover %>%
    separate(Stratum, into = c("Site", "Habitat", "Transect"), sep = "-")
total_turnover$Site_habitat <- paste(total_turnover$Site, total_turnover$Habitat)
# Add back in stratum column

# A lot of extra spaces

total_turnover <- total_turnover %>%
  mutate(across(everything(), str_trim))

str(total_turnover) # everything became characters

total_turnover$total <- as.numeric(total_turnover$total)
total_turnover$Year <- as.factor(total_turnover$Year)
total_turnover$Site <- as.factor(total_turnover$Site)
total_turnover$Habitat <- as.factor(total_turnover$Habitat)

turnover_means <-
  total_turnover %>%
  dplyr::group_by(Year, Site, Habitat, Site_habitat) %>%
  summarise(Turnover_mean = mean(total), 
            Turnover_se = sd(total)/sqrt(n()))

(turnover_plot <-
  ggplot(data = turnover_means) +
  geom_point(aes(x = Year, y = Turnover_mean, colour = Site, shape = Habitat, group = Site_habitat), size = 8, stroke = 2) +
  geom_line(aes(x = Year, y = Turnover_mean, colour = Site, group = Site_habitat), size = 1.2) +
  geom_errorbar(aes(x = Year, ymin = Turnover_mean - Turnover_se , ymax = Turnover_mean + Turnover_se, colour = Site,
                    group = Site_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = site_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site_habitat, nrow = 6)+
  labs(title = "a. Turnover\n", y = "Total turnover\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )
# this is useless

# rank shift

rank_shift<-
  rank_shift(
  df = transect_level,
  time.var = "Year",
  species.var = "Taxa",
  abundance.var = "Mean_cover",
  replicate.var = "Stratum"
)

rank_shift<- 
  rank_shift %>%
    separate(Stratum, into = c("Site", "Habitat", "Transect"), sep = "-")
rank_shift$Site_habitat <- paste(rank_shift$Site, rank_shift$Habitat)
# Add back in stratum column

# A lot of extra spaces

rank_shift <- rank_shift %>%
  mutate(across(everything(), str_trim))

str(rank_shift) # everything became characters

rank_shift$MRS <- as.numeric(rank_shift$MRS)
# Create a column with the final year from the returned time.var_pair
rank_shift$Year <- as.numeric(substr(rank_shift$year_pair, 6, 9))
rank_shift$Site <- as.factor(rank_shift$Site)
rank_shift$Habitat <- as.factor(rank_shift$Habitat)

rank_shift_means <-
  rank_shift %>%
  dplyr::group_by(Year, Site, Habitat, Site_habitat) %>%
  summarise(Rank_shift_mean = mean(MRS), 
            Rank_shift_se = sd(MRS)/sqrt(n()))

(rank_shift_plot <-
  ggplot(data = rank_shift_means) +
  geom_point(aes(x = Year, y = Rank_shift_mean, colour = Site, shape = Habitat, group = Site_habitat), size = 8, stroke = 2) +
  geom_line(aes(x = Year, y = Rank_shift_mean, colour = Site, group = Site_habitat), size = 1.2) +
  geom_errorbar(aes(x = Year, ymin = Rank_shift_mean - Rank_shift_se , ymax = Rank_shift_mean + Rank_shift_se, colour = Site,
                    group = Site_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = site_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site_habitat, nrow = 6)+
  labs(title = "a. rank_shift\n", y = "Total rank_shift\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )

# rank shift

rate_change<-
  rate_change_interval(
  df = transect_level,
  time.var = "Year",
  species.var = "Taxa",
  abundance.var = "Mean_cover",
  replicate.var = "Stratum"
)

rate_change<- 
  rate_change %>%
    separate(Stratum, into = c("Site", "Habitat", "Transect"), sep = "-")
rate_change$Site_habitat <- paste(rate_change$Site, rate_change$Habitat)

# A lot of extra spaces

rate_change <- rate_change %>%
  mutate(across(everything(), str_trim))

str(rate_change) # everything became characters

rate_change$distance <- as.numeric(rate_change$distance)
rate_change$interval <- as.numeric(rate_change$interval)
rate_change$Site <- as.factor(rate_change$Site)
rate_change$Habitat <- as.factor(rate_change$Habitat)

ggplot(rate_change, aes(interval, distance, colour = Site, shape = Habitat, group = Site_habitat, linetype = Habitat)) +
  facet_wrap(~Site) + 
  geom_point() +
  scale_colour_manual(values = site_colours) +
  scale_linetype_manual(values = habitat_linetypes) +
  stat_smooth(method = "lm", se = T, size = 2) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site)+
#  labs(y = "Distance\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
```
